Efficient ancestry and mutation simulation with msprime 1.0

### 2021-08-18 paper.tex

The \texttt{msprime.log\_arg\_likelihood(ts, r, N)} function
returns the natural logarithm of
the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
per-generation recombination probability \texttt{r} and population size \texttt{N}.
Specifically, the function returns the logarithm of
\begin{equation*}
\Bigg( \frac{ 1 }{ 2 N } \Bigg)^{ q_c } \Bigg( \prod_{ i : \mathcal{R} } r g_i \Bigg)
	\exp\Bigg( -\sum_{ i = 1 }^q \Big[\frac{ 1 }{ 2 N } \binom{ k_i }{ 2 }
		+ r l_i \Big] t_i  \Bigg),
\end{equation*}
where $t_i$ is the number of generations between the $(i - 1)$th and $i$th event,
$k_i$ is the number of extant ancestors in that interval, $l_i$ is the number of links
in that interval that would split ancestral material should they recombine,
$q$ is the total number of events in the tree sequence $\texttt{ts}$,
$q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices
of time intervals which end in a recombination, and $g_i$ is the corresponding
\emph{gap}:  the length of material in which the recombination at the end of the
$i$th period could have taken place without affecting the resulting genealogy.
A recombination in ancestral material will always have $g_i = 1$, while a recombination
in trapped non-ancestral material can result in other values of $g_i$.
For a continuous model of the genome and a recombination in ancestral material,
we set $g_i = 1$ and interpret the result as a density.

>>>>

The \texttt{msprime.log\_arg\_likelihood(ts, r, N)} function
returns the natural logarithm of
the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
per-generation recombination probability \texttt{r} and population size \texttt{N} \citep[e.g.][equation (1)]{kuhner2000maximum}.
Specifically, the function returns the logarithm of
\begin{equation*}
\Bigg( \frac{ 1 }{ 2 N } \Bigg)^{ q_c } \Bigg( \prod_{ i : \mathcal{R} } r g_i \Bigg)
	\exp\Bigg( -\sum_{ i = 1 }^q \Big[\frac{ 1 }{ 2 N } \binom{ k_i }{ 2 }
		+ r l_i \Big] t_i  \Bigg),
\end{equation*}
where $t_i$ is the number of generations between the $(i - 1)$th and $i$th event,
$k_i$ is the number of extant ancestors in that interval, $l_i$ is the number of links
in that interval that would split ancestral material should they recombine,
$q$ is the total number of events in the tree sequence $\texttt{ts}$,
$q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices
of time intervals which end in a recombination, and $g_i$ is the corresponding \emph{gap}:
the length of contiguous non-ancestral material around the link at which
the recombination in question took place. The gap indicates the number of links
(or length of genome in a continuous model) at which a recombination would result
in exactly the observed pattern of ancestral material in the ARG.
For a continuous model of the genome and a recombination in ancestral material,
we set $g_i = 1$ and interpret the result as a density.
